Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU

### 2016-01-26 paper/paperSource.tex
paper changes

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems: a collagen fiber structure with 890,000 atoms (top) and a dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system with 402,400 atoms (bottom) \cite{mainArticle} }
 \label{fg:collagen_dipal}
\end{figure}

@@

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/speedup838K-pomalo-za-4-eps-converted-to.pdf} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms \cite{mainArticle} }
 \label{fg:sample_estimation_old_paper}
\end{figure}

@@

Reading the data from files is a huge overhead, thus, speedup of push-based system in general is pretty significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads \cite{mainPaper}. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is very noticable how the framework \cite{mainPaper} contributes to efficiency of analysis of simulation data. You can observe some estimation samples in Figure \ref{fg:sample_estimation_old_paper}

On the other hand, some of the queries may be very slow due to their nature on a sequential type of computation,  especially, 2-body functions. Since the nature of the data that comes with simulation is basically physical properties of atoms, thus, there is a lot of computation that involves independent primitive mathematical operations, which makes it a perfect problem for parallelism.

Considering the fact that the running time of the most queries used
in the scientific world to analyze the MS is minuscule in comparison
to the loading time of the MS data, it is safe to say that the
scientific world would benefit from a system that can remedy the
aforementioned problem. Our idea is to build a system that would
load the MS data from hard disk in the main memory only once and
then execute as many queries as needed/wanted on that data, without
the need of reloading the same data over and over again. Considering
the volume of data in a single MS frame as well as the number of
frames produced during a single simulation, this type of system can
save a lot of time for MS analysis. The system would contain many,
most often used queries that can also be run as user's input. These
queries would be pre-programmed as separate modules in our system,
but they would be able to take certain attributes, like data
selection for instance, as user's input. Once the system is run, all
(or a selection) of these queries would be executed against the MS
data or certain selection. The system would act as a type of
push-based system, essentially pushing the loaded data to all the
queries. Once we have the data in the memory, the reading can loop
as many times as the queries need it. This is useful for some
non-streaming functions/queries that cannot be computed in a single
run of the data. Such one query is the SDH. But we have also come up
with a way to resolve this in a way, by creating a quad-tree based
database for this query. This quad-tree (we called Density Map (DM))
is being populated while the data is being loaded into main memory.
After the DM is populated, we can compute the SDH with much faster
algorithm than the brute-force method used everywhere else.

Another improvement that our system provides is the following: by
thorough observation and analysis we have discovered that many of
the quantities (functions/queries) that are being computed in order
to analyze the MS system share some of the same basic parts. In
order to take advantage of that, our systems precomputes these basic
parts and has them handy (stored in memory) whenever one of the more
complex functions needs them. So, for instance, when the total mass
of the system is needed, our system just pulls the precomputed value
out and serves the query that needs it immediately. And if more than
one query uses the same basic part, that translates into time
saving, however minuscule it might be.

\subsection{Contribution and roadmap of the paper} We have designed
and implemented a push-based system that can be used as a tool for
analysis of MS systems. We have tested the system on real MS
datasets. The experimental results evidently show the superiority of
the proposed system over one of the most widely used MS analysis
system across many scientific fields, i.e., GROMACS. The efficiency
improvement ranges anywhere from $3$ to $100$ times improvement (of
the overall running time) per set of selected queries run on a
different selection of the MS data. If we take in account the number
of such queries that the scientist run every day, we believe our
system would be of great benefit to MS community. In addition to the
higher efficiency that our system achieves, we also believe that our
idea will initiate a new breed of database related, push-based
systems that could be used to analyze the MS systems in a more
efficient way. Having a push-based like system, we were also able to
improve the computation of the more complex quantities. Namely, the
computation of some such complex quantities involves computation of
much less complex functions (sub-parts), and many such sub-parts are
being shared among some of the complex quantities. So, by computing
these more basic, sub-quantities in advance, we were able to achieve
however minuscule time gain when computing the more complex queries.

>>>>

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems: a collagen fiber structure with 890,000 atoms (top) and a dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system with 402,400 atoms (bottom) \cite{mainArticle}}
 \label{fg:collagen_dipal}
\end{figure}

@@

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/speedup838K-pomalo-za-4-eps-converted-to.pdf} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms \cite{mainPaper}}
 \label{fg:sample_estimation_old_paper}
\end{figure}

@@

Reading the data from files is a huge overhead, thus, speedup of push-based system in general is significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is quite noticeable how the framework contributes to efficiency of analysis of simulation data. You can observe some estimation samples in Figure \ref{fg:sample_estimation_old_paper}. Even though the memory loading and pushing steps are the key features of the proposed design, nevertheless, having general knowledge about the network of queries, in this paper, we will try to focus on optimizing the actual execution of them.

Some of the queries may be very slow due to their nature on a sequential type of computation,  especially, 2-body functions. Since the nature of the data that comes with simulation is basically physical properties of atoms, there is a lot of computation that involves independent primitive mathematical operations, which makes it a perfect problem for parallelism.

Although the original idea of push-based system for molecular simulation data analysis was focused on optimizing throughput and usage of loaded data, there were some extra approaches specifically for 2-body functions. For example, Density Map for Spacial Distance Histogram was used in order to avoid additional memory allocation and latency reduction with proven error bounds. SDH is a quite intensive and computational problem, especially with increasing amount of atoms.

We believe that having this nature of computational problems, the proposed GPU improved version of push-based system will significantly change in terms of performance by incorporating parallelism with CUDA.

\subsection{Contribution and roadmap of the paper} 
The proposed improved version of push-based system for molecular simulation data analysis, we believe, gives an opportunity for scientists to run their analysis even faster now. Taking an advantage of GPU devices nature and incorporating parallelism and streaming for different types of queries, we have come up with a good speed up over sequential processing, which already had a good speed up in terms of performance in comparison with original pull-based approach. Since in this paper we don't expect certain form of data, we have come up with a tool that generates mock data for any number of atoms and frames, which can be used for performance tests that were done consequently, too. This mock data has exactly the same information that would come with Gromacs simulation files, and it has exactly the same format as in the previous paper \cite{mainPaper} right before loading it to memory. This improved version has been developed on Amazon Elastic GPU nodes that consequently can be replicated and scaled up becoming more of a streaming processing engine, which can be very effective on costs, since it is always easy to scale up the nodes, shut them down and spin them back up as well as on a simple desktop computer with a GPU device.
